WASP-4 is Accelerating Towards the Earth

### 2020-02-03 paper/ms.tex
minor paper edits

\begin{abstract}
  Space and ground-based transit observations of the hot Jupiter
  WASP-4b have recently shown that its orbital period appears to be
  decreasing.  Proposed explanations for the period change include
  tidal orbital decay, orbital precession, and light-travel time
  effects.  We present new radial velocity measurements of WASP-4,
  acquired with Keck-HIRES.  The data show that the system is
  accelerating towards the Earth at $\dot{\gamma} =
  -0.0422^{+0.0028}_{-0.0027}\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.
  The implied period decrease explains most or all of WASP-4's
  changing orbital period as a light-travel time effect ($-5.94 \pm
  0.39~{\rm msec}\,{\rm yr}^{-1}$ implied, compared to $-8.64 \pm
  1.26~{\rm msec}\,{\rm yr}^{-1}$ observed).  Combining the radial
  velocities with new upper limits from speckle imaging, we find that
  the system's acceleration is likely caused by a 10-200$\,M_{\rm
  Jup}$ companion with semi-major axis between 10-100$\,$AU.  The
  statistics from \citet{knutson_friends_2014} imply that about 1 in 6
  hot Jupiters are expected to show comparable period decreases to
  WASP-4, due to acceleration by outer companions.  We find that these
  period decreases will be increasingly measureable in coming years,
  because the precision of period derivative measurement scales as
  (observing baseline)$^{5/2}$.  Continued radial velocity monitoring
  of hot Jupiters is therefore essential to distinguish tidal orbital
  decay from line-of-sight accelerations.
\end{abstract}

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f1.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Timing residuals and best-fit models for WASP-4b.}
  The vertical axis shows the observed transit times minus the
  calculated times assuming a constant orbital period.  More opaque
  points correspond to more precise data.  The $\pm1\sigma$
  uncertainties of the quadratic ephemeris 
  are shown in blue.
  The binned TESS point (yellow star) is the weighted average of 18
  TESS transits and is binned for display purposes only.  Its
  uncertainty is slightly larger than the marker size.  The models
  were fitted to all of the individual transit times.
  \label{fig:times}
	}
\end{figure}

@@

The resulting best-fit period derivative for the quadratic model is 
\begin{equation}
\dot{P}
  = - (2.74 \pm 0.28)\times 10^{-10}
  = - 8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees to within $1\sigma$ of the value reported by
\citet{southworth_transit_2019} ($\dot{P} = - 9.2 \pm 1.1~{\rm
ms}\,{\rm yr}^{-1}$).  It is $\approx$2.3$\sigma$ larger than the rate
of period decrease we previously reported \citepalias[$- 12.6 \pm 1.2~{\rm
ms}\,{\rm yr}^{-1}$;][]{bouma_wasp4b_2019}, presumably because of the
new data from \citeauthor{southworth_transit_2019} and
\citeauthor{baluev_2019}  The other best-fit transit timing model
parameters are reported in Table~3.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocity residuals of WASP-4.} The best-fit
    Keplerian orbit of WASP-4b has been subtracted.  The linear trend
    inferred from the RV data is shown with a black line, with
    $1\sigma$ errors in gray.  The trend that would be needed to
    produce the period decrease seen in transits ($\dot{P}= - 8.64 \pm
    1.26~{\rm msec}\,{\rm yr}^{-1}$) is indicated with the purple dotted
    line.  The four new RV measurements presented in this work
    increased the significance of the radial velocity trend from
    $\approx$2$\sigma$ to 15$\sigma$.
	\label{fig:rvs}
  \vspace{-0.3cm}
	}
\end{figure}

@@

In the preferred ``planet+trend'' model, WASP-4 is accelerating
towards our line-of-sight at high confidence,
\begin{equation}
  \dot{v}_{\rm r} = \dot{\gamma} = 
     -0.0422^{+0.0028}_{-0.0027}
     \,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}.
\end{equation}
For comparison, before our new measurements, $\dot{\gamma}$ was
thought to be about five times smaller, and had marginal statistical
significance \citep{knutson_friends_2014,bouma_wasp4b_2019}.

The system's acceleration towards our line-of-sight causes a decrease
in the apparent orbital period:
\begin{align}
  \dot{P}_{\rm\,RV} &= \frac{\dot{v}_{\rm r} P}{c},
\end{align}
or in more convenient units,
\begin{align}
  \dot{P}_{\rm\,RV} &= 105{.}3\,{\rm msec}\,{\rm yr}^{-1} \
  \left( \frac{P}{{\rm day}} \right)
  \left( \frac{\dot{\gamma}}{ {\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}} \right).
\end{align}
For WASP-4, this yields
\begin{align}
  \dot{P}_{\rm\,RV} &= -5.94 \pm 0.39\,{\rm msec}\,{\rm yr}^{-1}.
\end{align}
The majority of the period decrease observed in transits ($\dot{P}= -
8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$) seems to be explained by the
acceleration of the host star.

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f3.pdf}
	\end{center}
	\vspace{-0.7cm}
    \caption{
      {\bf Zorro contrast limits derived from point-source
      injection-recovery experiments.} Sources below the curve would
      have been detected.  The inset shows the speckle image
      reconstructed from 1000 40 millisecond frames in an 832$\,$nm
      bandpass, and acquired on September 28, 2019.  The image scale
      is $2.46''\times2.46''$.
    }
    \label{fig:zorro}
\end{figure}

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.78\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
    {\bf Masses and semi-major axes of companions that meet
    requirements of both the radial velocities and the speckle
    imaging.} The likelihood inferred from radial velocities is shown
    in grayscale, and the region excluded from the speckle imaging is
    indicated.  The expected mass to semimajor-axis degeneracy is
    shown with a black line.
	\label{fig:mass_sma}
  \vspace{-0cm}
	}
\end{figure*}

@@

\subsection{Implications for WASP-4}
Previous potential explanations for WASP-4b's decreasing orbital
period included tidal orbital decay, orbital precession, and
light-travel time effects \citep{bouma_wasp4b_2019}.  Our new radial
velocity measurements strongly indicate that the least exotic
option---light-travel time effects---is also the most likely.
Transits show the obital period decreasing by $- 8.64 \pm 1.26~{\rm
ms}\,{\rm yr}^{-1}$; the line-of-sight acceleration observed in radial
velocities would predict a period decrease of $-5.94 \pm 0.39~{\rm
ms}\,{\rm yr}^{-1}$.  Though the quantitative agreement is not
perfect, Occam's razor would suggest that most or all of the apparent
decrease of WASP-4b's orbital period is caused by the line-of-sight
acceleration.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.48\textwidth]{f5.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
  {\bf Predicted hot Jupiter period changes from linear radial
  velocity trends.} Including WASP-4b, 16 of 51 hot Jupiters from
  \citet{knutson_friends_2014} have shown long-term radial velocity
  trends.  HAT-P-11 signal is plotted, though its signal may be caused
  by stellar activity.  Three hot Jupiters are not plotted because
  their radial velocity curves better described as quadratic trends in
  time: HAT-P-17, WASP-8, and WASP-34.  Objects are ordered in the $y$
  dimension by the absolute value of d$P$/d$t$.
	\label{fig:pdot_pop}
  \vspace{-0.3cm}
	}
\end{figure}

@@

Including WASP-4b, 16 of 51 hot Jupiters surveyed by
\citet{knutson_friends_2014} show a non-zero radial velocity trend.
Therefore around 1 in 3 hot Jupiters is expected to show period
changes commensurate with WASP-4 due to acceleration by outer
companions.  Half of these will be period decreases\footnote{ At a
population level, one would expect half of the line-of-sight
accelerations to be positive, and half negative.  Tidal orbital decay
however is usually negative, and might manifest as an asymmetry in the
$\dot{P}$ distribution of hot Jupiters.}, and will be an
astrophysical false positive in searches for tidal orbital decay.

@@

For hot Jupiters that have been monitored for over baselines of
$\approx$10 years, secular changes in their orbital periods are
currently being constrained to a precision of $\lesssim 10~{\rm
ms}\,{\rm yr}^{-1}$ ({\it e.g.}, K.~Patra et al{.} 2020, submitted).
This is roughly commensurate with the level of signal many outer
companions are expected to induce (Figure~\ref{fig:pdot_pop}).

It is therefore pertinent to ask at what point in time further
detections of the light-travel time effect will become routine for hot
Jupiters.  This question is the same as asking: at what rate does the
uncertainty in the quadratic term of Equation~\ref{eq:quadratic_model}
scale with the observing baseline?  We take a Fisher analysis approach
to the problem.  First, we rewrite Equation~\ref{eq:quadratic_model}
as
\begin{align}
  t_{\rm tra} = a_0 + a_1 E + a_2 E^2,
\end{align}
where $a_0\equiv t_0$, $a_1\equiv P$, and $a_2\equiv 0.5\cdot {\rm
d}P/{\rm d}E$.  Following \citet{gould_chi2_2003}, one can show that
if $N$ transit timing measurements are taken uniformly across a
baseline of $\Delta E$ epochs with constant precision $\sigma_{\rm
tra}$, then the uncertainty of the quadratic term is given by
\begin{align}
  \sigma_{a_2} = 6\sqrt{5}
   \frac{\sigma}{N^{1/2} (\Delta E)^2} \propto (\Delta E)^{-5/2}.
\end{align}
This result implies that a doubled observing baseline yields an
$\approx$5.7-fold improvement in precision on d$P$/d$t$.  If regular
observations continue from ground and space-based observatories,
period derivatives will begin to be measured with precision exceeding
$1~{\rm msec}\,{\rm yr}^{-1}$ over the coming decade.

@@

Most hot Jupiters have outer companions with masses larger than
Jupiter beyond 5$\,$AU
\citep{knutson_friends_2014,bryan_statistics_2016}. The accelerations
and period changes induced by these outer companions will become an
increasingly large nuisance in the hunt for orbital decay as the
observational baselines get longer.  In particular, the precision with
which the period derivative can be measured from transits scales with
the baseline duration to the power of 5/2 (Section~\ref{sec:fisher}), and
so within a decade many more hot Jupiters should show orbital period
changes due to accelerations from their outer companions.  
To distinguish this effect from tidal decay, further long-term radial
velocity measurements of hot Jupiters are required.

>>>>

\begin{abstract}
  Space and ground-based transit measurements of the hot Jupiter
  WASP-4b have recently shown that its orbital period appears to be
  decreasing.  Proposed explanations for the period change include
  tidal orbital decay, orbital precession, and light-travel time
  effects.  We present new radial velocity measurements of WASP-4,
  acquired with Keck-HIRES.  The data show that the system is
  accelerating towards the Earth at $\dot{\gamma} =
  -0.0422^{+0.0028}_{-0.0027}\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.
  The implied period decrease explains most or all of WASP-4's
  changing orbital period as a light-travel time effect ($-5.94 \pm
  0.39~{\rm msec}\,{\rm yr}^{-1}$ implied, compared to $-8.64 \pm
  1.26~{\rm msec}\,{\rm yr}^{-1}$ observed).  Combining the radial
  velocities with new upper limits from speckle imaging, we find that
  the system's acceleration is likely caused by a 10-200$\,M_{\rm
  Jup}$ companion with semi-major axis between 10-100$\,$AU.  The
  statistics from \citet{knutson_friends_2014} imply that about 1 in 6
  hot Jupiters are expected to show comparable period decreases to
  WASP-4, due to acceleration by outer companions.  These
  period decreases will be increasingly measureable in coming years,
  because the precision of the period derivative measurement scales with
  the observing baseline to the power of 5/2.  Continued radial velocity monitoring
  of hot Jupiters is therefore essential to distinguish tidal orbital
  decay from line-of-sight accelerations.
\end{abstract}

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.47\textwidth]{f1.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Timing residuals and best-fit models for WASP-4b.}
  The vertical axis shows the observed transit times minus the
  calculated times assuming a constant orbital period.  More opaque
  points correspond to more precise data.  The $\pm1\sigma$
  uncertainties of the quadratic ephemeris 
  are shown in blue.
  The binned TESS point (yellow star) is the weighted average of 18
  TESS transits and is binned for display purposes only.  The models
  were fitted to all of the individual transit times.
  \label{fig:times}
	}
\end{figure}

@@

The resulting best-fit period derivative for the quadratic model is 
\begin{equation}
\dot{P}
  = - (2.74 \pm 0.28)\times 10^{-10}
  = - 8w.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees to within $1\sigma$ of the value reported by
\citet{southworth_transit_2019} ($\dot{P} = - 9.2 \pm 1.1~{\rm
ms}\,{\rm yr}^{-1}$).  It is $\approx$2.3$\sigma$ larger than the rate
of period decrease we previously reported \citepalias[$- 12.6 \pm 1.2~{\rm
ms}\,{\rm yr}^{-1}$;][]{bouma_wasp4b_2019}, presumably because of the
new data from \citeauthor{southworth_transit_2019} and
\citeauthor{baluev_2019}  The other best-fit transit timing model
parameters are reported in Table~3.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.47\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocity residuals of WASP-4.} The best-fit
    Keplerian orbit of WASP-4b has been subtracted.  The linear trend
    inferred from the RV data is shown with a black line, with
    $1\sigma$ errors in gray.  The trend that would be needed to
    produce the period decrease seen in transits ($\dot{P}= - 8.64 \pm
    1.26~{\rm msec}\,{\rm yr}^{-1}$) is indicated in purple.  The four
    new RV measurements presented in this work increase the
    significance of the radial velocity trend from $\approx$2$\sigma$
    to 15$\sigma$.
	\label{fig:rvs}
  \vspace{-0.3cm}
	}
\end{figure}

@@

In the preferred ``planet+linear trend'' model, WASP-4 is accelerating
towards our line-of-sight at high confidence,
\begin{equation}
  \dot{v}_{\rm r} = \dot{\gamma} = 
     -0.0422^{+0.0028}_{-0.0027}
     \,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}.
\end{equation}
For comparison, before our new measurements, $\dot{\gamma}$ was
thought to be about five times smaller, and had marginal statistical
significance \citep{knutson_friends_2014,bouma_wasp4b_2019}.

The system's acceleration towards our line-of-sight causes a decrease
in the apparent orbital period:
\begin{align}
  \dot{P}_{\rm\,RV} &= \frac{\dot{v}_{\rm r} P}{c},
\end{align}
or in more convenient units,
\begin{align}
  \dot{P}_{\rm\,RV} &= 105{.}3\,{\rm msec}\,{\rm yr}^{-1} \
  \left( \frac{P}{{\rm day}} \right)
  \left( \frac{\dot{\gamma}}{ {\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}} \right).
\end{align}
For WASP-4, this yields
\begin{align}
  \dot{P}_{\rm\,RV} &= -5.94 \pm 0.39\,{\rm msec}\,{\rm yr}^{-1}.
\end{align}
The majority of the period decrease seen in transits ($\dot{P}= -
8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$) therefore seems to be caused by the
acceleration of the host star.

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.47\textwidth]{f3.pdf}
	\end{center}
	\vspace{-0.7cm}
    \caption{
      {\bf Zorro contrast limits derived from point-source
      injection-recovery experiments.} Sources below the curve would
      have been detected.  The inset shows the speckle image
      reconstructed from 1000 60 millisecond frames in an 832$\,$nm
      bandpass, and acquired on September 28, 2019.  The image scale
      is $2.46''\times2.46''$.
    }
    \label{fig:zorro}
\end{figure}

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.73\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
    {\bf Masses and semi-major axes of companions that meet
    requirements of both the radial velocities and the speckle
    imaging.} The likelihood inferred from radial velocities is shown
    in grayscale, and the region excluded from the speckle imaging is
    indicated.  The expected mass to semimajor-axis degeneracy is
    shown with a black line.
	\label{fig:mass_sma}
  \vspace{-0cm}
	}
\end{figure*}

@@

\subsection{Implications for WASP-4}
Previous potential explanations for WASP-4b's decreasing orbital
period included tidal orbital decay, orbital precession, and
light-travel time effects \citep{bouma_wasp4b_2019}.  Our new radial
velocity measurements strongly indicate that the least exotic
option---light-travel time effects---is also the most likely.
Transits show the obital period decreasing by $- 8.64 \pm 1.26~{\rm
ms}\,{\rm yr}^{-1}$; the line-of-sight acceleration observed in radial
velocities would predict a period decrease of $-5.94 \pm 0.39~{\rm
ms}\,{\rm yr}^{-1}$.  Though the quantitative agreement is $\approx$2$\sigma$
discrepant, Occam's razor would suggest that most or all of the apparent
decrease of WASP-4b's orbital period is caused by the line-of-sight
acceleration.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.48\textwidth]{f5.pdf}
	\end{center}
	\vspace{-0.4cm}
	\caption{
  {\bf Predicted hot Jupiter period changes from linear radial
  velocity trends.} Including WASP-4b, 16 of 51 hot Jupiters from
  \citet{knutson_friends_2014} have shown long-term radial velocity
  trends.  HAT-P-11 is shown, though its signal may be caused
  by stellar activity.  Three hot Jupiters are not shown because
  their radial velocity curves are better described as quadratic
  trends in time: HAT-P-17, WASP-8, and WASP-34.  Objects are ordered
  in the $y$ dimension by the absolute value of d$P$/d$t$.
	\label{fig:pdot_pop}
  \vspace{-0.3cm}
	}
\end{figure}

@@

Including WASP-4b, 16 of 51 hot Jupiters surveyed by
\citet{knutson_friends_2014} show a non-zero radial velocity trend.
Therefore around 1 in 3 hot Jupiters is expected to show period
changes commensurate with WASP-4 due to acceleration by outer
companions.  Half of these will be period decreases\footnote{ At a
population level, one would expect half of the line-of-sight
accelerations to be positive, and half negative.  Tidal orbital decay
however is usually negative, and could evenutally manifest as an asymmetry in the
$\dot{P}$ distribution of hot Jupiters.}, and will be an
astrophysical false positive in searches for tidal orbital decay.

@@

For hot Jupiters that have been monitored over baselines
exceeding 10 years, secular changes in their orbital periods are
currently being constrained to a precision of $\lesssim$$10~{\rm
msec}\,{\rm yr}^{-1}$ ({\it e.g.}, K.~Patra et al{.} 2020, submitted).
This is roughly commensurate with the level of signal many outer
companions are expected to induce (Figure~\ref{fig:pdot_pop}).

It is therefore pertinent to ask at what point in time further
detections of the light-travel time effect will become routine for hot
Jupiters.  This question is the same as asking: at what rate does the
uncertainty in the quadratic term of Equation~\ref{eq:quadratic_model}
scale with the observing baseline?  We take a Fisher analysis approach
to the problem.  First, we rewrite Equation~\ref{eq:quadratic_model}
as
\begin{align}
  t_{\rm tra} = a_0 + a_1 E + a_2 E^2,
\end{align}
where $a_0\equiv t_0$, $a_1\equiv P$, and $a_2\equiv 0.5\cdot {\rm
d}P/{\rm d}E$.  Following \citet{gould_chi2_2003}, one can show that
if $N$ transit timing measurements are taken uniformly across a
baseline of $\Delta E$ epochs with constant precision $\sigma$, then
the uncertainty of the quadratic term is given by
\begin{align}
  \sigma_{a_2} = 6\sqrt{5}
   \frac{\sigma}{N^{1/2} (\Delta E)^2} \propto (\Delta E)^{-5/2}.
\end{align}
This result implies that a doubled observing baseline yields an
$\approx$5.7-fold improvement in precision on d$P$/d$t$.  If regular
observations continue from ground and space-based observatories,
period derivatives will be measured with precision exceeding
$1~{\rm msec}\,{\rm yr}^{-1}$ within the coming decade.

@@

Most hot Jupiters have outer companions with masses larger than
Jupiter beyond 5$\,$AU
\citep{knutson_friends_2014,bryan_statistics_2016}. The accelerations
and period changes induced by these outer companions will become an
increasingly large nuisance in the hunt for tidal orbital decay as the
observational baselines get longer.  In particular, the precision with
which the period derivative can be measured from transits scales with
the baseline duration to the power of 5/2 (Section~\ref{sec:fisher}), and
so within a decade many more hot Jupiters should show orbital period
changes due to accelerations from their outer companions.  
To distinguish this effect from tidal decay, further long-term radial
velocity measurements of hot Jupiters are strongly encouraged.
